\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-07 paper/algorithm.tex
paper: formatting

\subsection{Cost model}
\label{sec:cost-model}
We rank query plans $Q$ using a cost model $\text{Cost}(Q) = (1 - \alpha) \times \text{Time}(Q) + \alpha \times \text{Loss}(Q)$. $\text{Time}(Q)$ is an estimate of the runtime of the query which is derived from table statistics and selectivity estimation of the query predicates. $\text{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure. $\alpha$ is a parameter to the query optimizer that controls the emphasis on quality over performance. $\alpha = 1.0$ means that the query should be as accurate as possible (at the expense of performance), $\alpha=0.0$ means that the query should be as fast
as possible (at the expense of accuracy).

In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

For each sub-query, we keep track of the estimated number of
null values in a column  ($\text{MissingCount}(c, Q)$) by using the histograms associated with
the sub-query. We use this estimate to compute
$\text{Loss}(Q)$ as follows.

\begin{align*}
  \text{Loss}(Q) = \begin{cases}
    \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \delta_C(Q') \\
    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \mu_C(Q') \\
    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
tuples available to train the imputation algorithm increases.  Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more complete values to work
with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
regression trees. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

>>>>

\subsection{Cost model}
\label{sec:cost-model}
We rank query plans using a cost model \[\textsc{Cost}(Q) = (1 - \alpha) \times \textsc{Time}(Q) + \alpha \times \textsc{Loss}(Q).\] $\textsc{Time}(Q)$ is an estimate of the runtime of the query derived from table statistics and selectivity estimation of the query predicates. $\textsc{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure. $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance. $\alpha = 1.0$ means that the optimal query should be as accurate as possible (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast
as possible (at the expense of accuracy).

In order to correctly estimate $\textsc{Time}(Q)$ and $\textsc{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

For each sub-query, we keep track of the estimated number of
null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
the sub-query. We use this estimate to compute
$\text{Loss}(Q)$ as follows.

\begin{align*}
  \textsc{Loss}(q) = \begin{cases}
    \sum_{c \in C} \textsc{Missing}(c, q') & q = \delta_C(q') \\
    \frac{1}{\sqrt{|q'|}} \sum_{c \in C} \textsc{Missing}(c, q') & q = \mu_C(q') \\
    \textsc{Loss}(q_1') + \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{Loss}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
tuples available to train the imputation algorithm increases.  Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more complete values to work
with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
regression trees. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.
